---
title: "Homework 6"
format: 
  html:
    self-contained: true
    embed-resources: true
editor: visual
execute: 
  echo: true
  include: true
---

For the problems in which calculations are needed, please include your R code with your answers, otherwise you will not be given full credit. Please upload your assignment by Wednesday, November 6 at 6 pm in a pdf/html/word file to Sakai.

Consider the awards.csv dataset in the Data folder that contains the following variables:

-   ID: Student identification

-   num_awards: Number of Awards the student has won

-   prog: The type of program the student is enrolled in

-   math: Their math proficiency score

    ```{r}
    library(tidyverse)
    award <- read.csv('Data/awards.csv')
    ```

1.  We wish to create a regression model to predict the number of awards a student has won by their program and math efficiency.

<!-- -->

a.  State the regression line for the distribution's natural parameter in the context of the problem.

    $log(\hat{awards}) = \beta_0 + \beta_1 General + \beta_2 Vocational + \beta3 Math$

```{r}
# using poisson bc awards is a count variable 
mod <- glm(num_awards ~ prog + math, family = poisson, data = award)
summary(mod)
```

b.  Interpret the parameter associated with math in the context of the problem.

    0.07015: As the math proficiency score of a student increases by one, the expected increase in the number of awards for the student changes by a factor of exp(0.07015) = 1.073, holding program constant.

c.  Calculate and interpret a 95% profile confidence interval for the parameter associated with the vocational program in the context of the problem.

```{r}
confint(mod, 'progVocational')
```

We are 95% confident that if the program a student is in is

d.  Perform a formal hypothesis test to determine if your model in (a) is statistically significant. Assume $\alpha = 0.05$.

```{r}
```

e.  Perform a formal hypothesis test to determine if adding interaction terms between program and math score is statistically significant. Assume $\alpha = 0.05$.

```{r}
```

f.  Determine if the model you chose in (e) is a good fit for the data.

```{r}

```
